Tight Generalization Bounds for Hard Parameter Sharing in High-dimensional Linear Regression

### 2020-10-06 arxiv/diff_cov.tex
section 5 v0 wip

\begin{example}[Covariate shift]\label{sec_covshift}
%So far we have considered the isotropic model where $\Sigma_1 = \Sigma_2$.
%This setting is relevant for settings where different tasks share the same input features such as multi-class image classification.
%In general, the covariance matrices of the two tasks may be different such as in text classification.
Our second example studies the impact of varying covariate shifts on the variance limit.
%As we are going to show later, covariate shift is accurately captured by the spectrum of $\Sigma^{1/2}\Sigma^{-1/2}$.
Recall that the matrix $M$ captures the geometry of covariate shifts.
Suppose that half of the singular values of $M$ are equal to a parameter $\lambda > 0$ and the other half are equal to $\lambda^{-1}$.
We observe the following dichotomy on the impact of covariate shifts.
\begin{itemize}
	\item If $n_1 \ge n_2$, then the variance (limit) of hard parameter sharing is smallest when $\lambda = 1$ or there is no covariate shift.
	\item If $n_1 < n_2$, then having covariate shifts ($\lambda \neq 1$) always helps reduce variance (limit).
\end{itemize}
To see this, we first write the variance limit for different values of $\lambda$ as follows
\begin{align*}
	\bigtr{\Sigma^{(2)} (a_1 \Sigma^{(1)} + a_2 \Sigma^{(2)})^{-1}}
	= \bigtr{(a_1 M^{\top} M + a_2 \id)^{-1}}
	= \frac{1}{2}\bigbrace{(\lambda^{-2} a_1 + a_2)^{-1} + (\lambda^2 a_1 + a_2)^{-1}}.
\end{align*}
Denote the above function by $f(\lambda)$.
We compare $f(\lambda)$ with $f(1)$, and observe that
\[ 2f(\lambda) - 2f(1) = {(\lambda^2 a_1 + \frac{n_1 + n_2 - p}{n_1 + n_2} - a_1})^{-1}
	+ (\lambda^{-2} a_1 + \frac{n_1 + n_2 - p}{n_1 + n_2} - a_1)^{-1}  - \frac{2(n_1 + n_2)}{n_1 + n_2 - p}, \]
where we used the first part of equation \eqref{lem_cov_shift_eq}.
Next, one can verify that the above equation is equal to $(2a_1 - \frac{n_1 + n_2 - p} {n_1 + n_2}) g(\lambda, a_1)$, for a fixed and positive function $g(\lambda, a_1)$ (details omitted).
Hence, whether or not $f(\lambda)$ is small than $f(1)$ is decided by $a_1$.
To solve $a_1$, we present an informal argument that $a_1 \ge \frac{(n_1 + n_2 - p)}{2(n_1 + n_2)}$ if and only if $n_1 \ge n_2$.
Intuitively, with the current spectrum of $M$, $a_1$ and $a_2$ have a symmetric role in the self-consistent equations.
Hence, when $n_1 \ge n_2$, we expect that $a_1 \ge a_2$, hence $a_1 \ge \frac{1}{2}(1 - \frac{p}{n_1 + n_2})$.
Vice serva when $n_1 < n_2$.
A formal proof requires solving the self-consistent equations \eqref{lem_cov_shift_eq} by replacing half of the singular with $\lambda$ and the other half with $\lambda^{-1}$.
We omit the details.
Thus, we conclude that if $n_1 \ge n_2$, then $f(\lambda) > f(1)$; if $n_1< n_2$, then $f(\lambda)< f(1)$.
\end{example}

@@

Our result shows that the prediction loss of hard parameter sharing consists of a bias term that scales with task-specific variance $d^2$ and a noise variance term that scales with $\sigma^2$.
In the following, we explain an interesting phenomenon, where adding task one's samples helps task two initially, but hurts eventually.
Consider the prediction loss if we fix $n_2$ and increase $n_1$.
First, the noise variance term in Corollary \ref{cor_MTL_loss} always reduces as $n_1$ increases.
Second, the bias term always increases as $n_1$ increases, which can be verified by calculating the derivative of the bias term.
Combined together, we observe that the \textit{optimal} sample ratio between $n_1$ and $n_2$ should be among one of following three cases:
(i) $n_1 = 0$, in which case the prediction loss reduces to $\frac{\sigma^2 p}{n_2 - p}$;
(ii) $n_1 = \infty$, in which case the prediction loss reduces to $d^2$;
(iii) the critial point where the derivative of the bias and the variance terms w.r.t. $n_1$ equals zero.
The proof can be found in Appendix \ref{app_iso_cov}.

Based on this observation, we have the following trend when we fix $n_2$ and increase $n_1$:
\begin{itemize}
	\item If $d^2$ is smaller than the prediction loss of STL for task two $\frac{\sigma^2 p}{n_2 - p}$, then the prediction loss of hard parameter sharing is always lower than STL. That is, task one always transfers positively to task two.
	\item Otherwise, task one transfers positively to task two initially, but transfers negatively eventually.
\end{itemize}
%Therefore, while the variance of MTL still reduces compared to STL,
%If the bias increases more than the amount of variance reduced, we observe negative transfer.
%of MTL increases and converges to $pd^2$ as the sample ratio $\rho_1/\rho_2$ increases, and the variance reduction $\frac{\sigma^2}{ \rho_2-1}  -\frac{\sigma^2}{\rho_1+\rho_2-1}$ also increases and converges to $\frac{\sigma^2}{\rho_2-1}$ as $\rho_1/\rho_2$ increases. The tradeoff between these two terms depends on $\rho_1$ in a nonlinear way that is more complicated than the one in the case of varying model distance. But from our results, one can check the following phenomena. If $\frac{\sigma^2}{\rho_2-1} > pd^2$, then the transfer is always positive. On the other hand, if $\frac{\sigma^2}{\rho_2-1} < pd^2$, then there is a transition from the positive transfer to negative transfer as $\rho_1/\rho_2$ increases.
%Therefore, while the variance of MTL still reduces compared to STL,
%If the bias increases more than the amount of variance reduced, we will observe negative transfer.
%Figure \ref{fig_size} provides a simulation result for such a setting (described in Appendix \ref{app_synthetic}).
%We observe that as $n_1 / n_2$ increases, there is a transition from positive to negative transfer.
%There are several unexplained observations near $y = 0$ caused by the concentration error on the right-hand side of \eqref{cor_MTL_error}.
\end{example}

%\begin{proposition}[Source/target sample ratio]\label{prop_data_size}
%	In the isotropic model, suppose that $\rho_1 > 40$ and $\rho_2 > 110$ are fixed constants, and $\Psi(\beta_1, \beta_2) > 2/(\rho_2 - 1)$.
%	Then we have that
%	\begin{itemize}
%		\item \textbf{Positive transfer:} If $\frac{n_1}{n_2} = \frac{\rho_1}{\rho_2} < \frac{1}{\nu} \cdot \frac{1 - 2\rho_2^{-1}}{\Psi(\beta_1, \beta_2) (\rho_2 - 1) - \nu^{-1}}$, then w.h.p. $$\te(\hat{\beta}_2^{\MTL}) < \te(\hat{\beta}_2^{\STL}).$$
%		\item \textbf{Negative transfer:} If $\frac{n_1}{n_2} = \frac{\rho_1}{\rho_2} > {\nu} \cdot \frac{1 - 2\rho_2^{-1}}{\Psi(\beta_1, \beta_2) (\rho_2 - 3/2) - \nu}$, then w.h.p. $$\te(\hat{\beta}_2^{\MTL}) > \te(\hat{\beta}_2^{\STL}).$$
%	\end{itemize}
%\end{proposition}
%Proposition \ref{prop_data_size} describes the bias-variance tradeoff in terms of the sample ratio $\rho_1 / \rho_2$.

>>>>

@@

Our result shows that the prediction loss of hard parameter sharing consists of a bias term that scales with task-specific variance $d^2$ and a noise variance term that scales with $\sigma^2$.
In the following, we explain an interesting phenomenon, where adding task one's samples helps task two initially, but hurts eventually.
Consider the prediction loss if we fix $n_2$ and increase $n_1$.
We show that as a function of $n_1$, the limit equation always decreases first from $\frac{\sigma^2 p}{n - p}$ with $n_1$ being zero, and then increases to $d^2$ with $n_1$ going to infinity.
We describe an informal proof.
First, the noise variance term in Corollary \ref{cor_MTL_loss} always reduces as $n_1$ increases.
Second, the bias term always increases as $n_1$ increases, which can be verified by calculating the derivative of the bias term.
Moreover, near the point where $n_1$ is zero, one can verify that bias increases less than the decrease in variance, and there is only one positive root for the derivative being zero.
%Combined together, we observe that the \textit{optimal} sample ratio between $n_1$ and $n_2$ should be among one of following three cases:
%(i) $n_1 = 0$, in which case the prediction loss reduces to $\frac{\sigma^2 p}{n_2 - p}$;
%(ii) $n_1 = \infty$, in which case the prediction loss reduces to $d^2$;
%(iii) the critial point where the derivative of the bias and the variance terms w.r.t. $n_1$ equals zero.
The proof of Corollary \ref{cor_MTL_loss} can be found in Appendix \ref{app_iso_cov}.

%Based on this observation, we have the following trend when we fix $n_2$ and increase $n_1$:
%\begin{itemize}
%	\item If $d^2$ is smaller than $\frac{\sigma^2 p}{n_2 - p}$, then the prediction loss of hard parameter sharing is always lower than STL. That is, task one always transfers positively to task two.
%	\item Otherwise, task one transfers positively to task two initially, but transfers negatively eventually.
%\end{itemize}
%Therefore, while the variance of MTL still reduces compared to STL,
%If the bias increases more than the amount of variance reduced, we observe negative transfer.
%of MTL increases and converges to $pd^2$ as the sample ratio $\rho_1/\rho_2$ increases, and the variance reduction $\frac{\sigma^2}{ \rho_2-1}  -\frac{\sigma^2}{\rho_1+\rho_2-1}$ also increases and converges to $\frac{\sigma^2}{\rho_2-1}$ as $\rho_1/\rho_2$ increases. The tradeoff between these two terms depends on $\rho_1$ in a nonlinear way that is more complicated than the one in the case of varying model distance. But from our results, one can check the following phenomena. If $\frac{\sigma^2}{\rho_2-1} > pd^2$, then the transfer is always positive. On the other hand, if $\frac{\sigma^2}{\rho_2-1} < pd^2$, then there is a transition from the positive transfer to negative transfer as $\rho_1/\rho_2$ increases.
%Therefore, while the variance of MTL still reduces compared to STL,
%If the bias increases more than the amount of variance reduced, we will observe negative transfer.
%Figure \ref{fig_size} provides a simulation result for such a setting (described in Appendix \ref{app_synthetic}).
%We observe that as $n_1 / n_2$ increases, there is a transition from positive to negative transfer.
%There are several unexplained observations near $y = 0$ caused by the concentration error on the right-hand side of \eqref{cor_MTL_error}.
\end{example}

\begin{example}[Covariate shift]\label{ex_covshift}
%So far we have considered the isotropic model where $\Sigma_1 = \Sigma_2$.
%This setting is relevant for settings where different tasks share the same input features such as multi-class image classification.
%In general, the covariance matrices of the two tasks may be different such as in text classification.
Our second example studies the impact of varying covariate shifts on the variance limit.
%As we are going to show later, covariate shift is accurately captured by the spectrum of $\Sigma^{1/2}\Sigma^{-1/2}$.
Recall that the matrix $M$ captures the geometry of covariate shifts.
Suppose that half of the singular values of $M$ are equal to a parameter $\lambda > 0$ and the other half are equal to $\lambda^{-1}$.
We observe the following dichotomy on the impact of covariate shifts.
\begin{itemize}
	\item If $n_1 \ge n_2$, then the variance (limit) of hard parameter sharing is smallest when $\lambda = 1$ or there is no covariate shift.
	\item If $n_1 < n_2$, then having covariate shifts ($\lambda \neq 1$) always helps reduce variance (limit).
\end{itemize}
To see this, we first write the variance limit for different values of $\lambda$ as follows
\begin{align*}
	\bigtr{\Sigma^{(2)} (a_1 \Sigma^{(1)} + a_2 \Sigma^{(2)})^{-1}}
	= \bigtr{(a_1 M^{\top} M + a_2 \id)^{-1}}
	= \frac{1}{2}\bigbrace{(\lambda^{-2} a_1 + a_2)^{-1} + (\lambda^2 a_1 + a_2)^{-1}}.
\end{align*}
Denote the above function by $f(\lambda)$.
We compare $f(\lambda)$ with $f(1)$, and observe that
\[ 2f(\lambda) - 2f(1) = {(\lambda^2 a_1 + \frac{n_1 + n_2 - p}{n_1 + n_2} - a_1})^{-1}
	+ (\lambda^{-2} a_1 + \frac{n_1 + n_2 - p}{n_1 + n_2} - a_1)^{-1}  - \frac{2(n_1 + n_2)}{n_1 + n_2 - p}, \]
where we used the first part of equation \eqref{lem_cov_shift_eq}.
Next, one can verify that the above equation is equal to $(2a_1 - \frac{n_1 + n_2 - p} {n_1 + n_2}) g(\lambda, a_1)$, for a fixed and positive function $g(\lambda, a_1)$ (details omitted).
Hence, whether or not $f(\lambda)$ is small than $f(1)$ is decided by $a_1$.
To solve $a_1$, we present an informal argument that $a_1 \ge \frac{(n_1 + n_2 - p)}{2(n_1 + n_2)}$ if and only if $n_1 \ge n_2$.
Intuitively, with the current spectrum of $M$, $a_1$ and $a_2$ have a symmetric role in the self-consistent equations.
Hence, when $n_1 \ge n_2$, we expect that $a_1 \ge a_2$, hence $a_1 \ge \frac{1}{2}(1 - \frac{p}{n_1 + n_2})$.
Vice serva when $n_1 < n_2$.
A formal proof requires solving the self-consistent equations \eqref{lem_cov_shift_eq} by replacing half of the singular with $\lambda$ and the other half with $\lambda^{-1}$.
We omit the details.
Thus, we conclude that if $n_1 \ge n_2$, then $f(\lambda) > f(1)$; if $n_1< n_2$, then $f(\lambda)< f(1)$.
\end{example}
%\begin{proposition}[Source/target sample ratio]\label{prop_data_size}
%	In the isotropic model, suppose that $\rho_1 > 40$ and $\rho_2 > 110$ are fixed constants, and $\Psi(\beta_1, \beta_2) > 2/(\rho_2 - 1)$.
%	Then we have that
%	\begin{itemize}
%		\item \textbf{Positive transfer:} If $\frac{n_1}{n_2} = \frac{\rho_1}{\rho_2} < \frac{1}{\nu} \cdot \frac{1 - 2\rho_2^{-1}}{\Psi(\beta_1, \beta_2) (\rho_2 - 1) - \nu^{-1}}$, then w.h.p. $$\te(\hat{\beta}_2^{\MTL}) < \te(\hat{\beta}_2^{\STL}).$$
%		\item \textbf{Negative transfer:} If $\frac{n_1}{n_2} = \frac{\rho_1}{\rho_2} > {\nu} \cdot \frac{1 - 2\rho_2^{-1}}{\Psi(\beta_1, \beta_2) (\rho_2 - 3/2) - \nu}$, then w.h.p. $$\te(\hat{\beta}_2^{\MTL}) > \te(\hat{\beta}_2^{\STL}).$$
%	\end{itemize}
%\end{proposition}
%Proposition \ref{prop_data_size} describes the bias-variance tradeoff in terms of the sample ratio $\rho_1 / \rho_2$.
